# Diagnostic horizons and other differentiae {#sec-diagnostics}

Diagnostic horizons and other differentiating criteria are defined to facilitate the assignment of soils to classes. The definitions are not intended to represent a comprehensive classification of horizons. A summary of these is given below.

```{r}
library(flextable)

t2 <- data.frame(
  'col1' = c('Horizons, Pans and\n Layers:', 'Argillic horizon', 'Brittle-B horizon',
             'Calcareous horizon', 'Cutanic horizon', 'Cutanoxidic horizon', 'Densipan',
             'Duripan', 'Distinct topsoil', 'Fragipan', 'Humus-pan', 'Ironstone-pan', 
             'Nodular horizon', 'Ortstein-pan', 'Oxidic horizon', 'Peaty topsoil',
             'Placic horizon', 'Podzolic-B horizon', 'Reductimorphic horizon',
             'Redox-mottled horizon', 'Slowly permeable layer', 'Weathered-B horizon'),
  'col2' = c('Soil materials and\n Contacts:', 'Allophanic soil material', 'Lithic contact',
             'Organic soil material', 'Paralithic contact', 'Tephric soil material',
             'Vitric soil material', '', 'Profile Forms:', 'Gley profile form', 'Mottled profile form',
             '', 'Features:', 'Fluvial features', 'Perch-gley features', 'Sodic features', '',
             'Other Differentiae:', 'Crumb structure', 'Reactive-aluminium test',
             'Redox segregations', ''))

ft <- flextable(t2)
ft <- delete_part(x = ft, part = "header")
ft <- bold(ft,  1, 1)
ft <- bold(ft,  1, 2)
ft <- bold(ft,  9, 2)
ft <- bold(ft, 13, 2)
ft <- bold(ft, 18, 2)
ft <- set_table_properties(ft, layout = "autofit", width = 1) 
ft
```

The following diagnostic horizons and soil characteristics have been modified from Soil Taxonomy @soilsurveystaff1999: [argillic horizon](#sec-diag-argh), [duripan](#sec-diag-dpan), [fragipan](#sec-diag-fpan), [lithic contact](#sec-diag-lithc), [paralithic contact](#sec-diag-plith) and [placic horizon](#sec-diag-plac).

## Allophanic Soil Material {#sec-diag-alloph}

Allophanic soil material has soil properties dominated by minerals with short-range order, especially allophane, imogolite and ferrihydrite. Other terms such as "amorphous", "poorly crystalline" and "non-crystalline", have been, or are, sometimes used for such minerals. Their chief characteristics are reactive variable-charge surfaces, and a very high specific surface area (several hundreds of m^2^/g). Allophanic soil material has **BOTH**

1.  ***Either***
    (a) All of the following (in a fresh sample):
        (i) Sensitive or strongly sensitive sensitivity class (distinctly greasy or smeary feel except in some sandy soils), *and*
        (ii) Very weak or weak unconfined soil strength class (when moist), *and*
        (iii) Non-sticky or slightly sticky stickiness class, *and*
        (iv) Strong or very strong reactive-aluminium test;\
             ***OR***
    (b) P retention of 85% or more;\
        **AND**
2.  Dry bulk density of the fine-earth fraction (where the volume is determined on a field-moist soil) of less than 0.9 Mg/m^3^.

Layers meeting the requirements of allophanic soil material may also meet the requirements of [vitric soil material](#sec-diag-vitr).

Accessory chemical properties relate to variable-charge characteristics, P retention, and high organic-matter contents. Accessory physical properties include high total available water capacity and readily available water capacity, and low penetration resistance. In addition, allophanic soil material undergoes irreversible changes upon drying, for example, in plastic and liquid limits and in apparent particle-size distribution. It should be noted that minerals other than allophane (e.g. ferrihydrite) can give rise to allophanic soil material.

## Argillic Horizon {#sec-diag-argh}

An argillic horizon is a clay-enriched horizon. It is indicated by a "Bt" horizon notation as in Btg, BCt, etc. It has **ONE** of the following:

1.  It is vertically continuous and is 10 cm or more thick. Clay coatings occur that have a waxy lustre when dry and sufficient thickness to envelop fine sand grains. Coatings occur either on peds (10% or more of the ped surfaces), or in pores (in more than one-third of the observed tubular pores) or as bridges between sand grains (in more than half of the horizon); **OR**
2.  It is composed of clay-enriched lamellae (clay bands), that within 90 cm of the mineral soil surface have a combined thickness of 15 cm or more; **OR**
3.  It contains sufficiently more clay than the overlying horizon, as detected by hand texturing (5% or more) , excluding any differences which result from a lithological discontinuity, and **either**
    (a) it is overlain by an eluvial horizon (Ew or Eg) and the upper boundary of texture contrast is abrupt or sharp, **or**
    (b) clay coatings occur on ped or pore surfaces.

Horizons having coatings which do not meet the requirements of an argillic horizon are likely to meet the requirements of a [cutanic horizon](#sec-diag-cuth).

## Brittle-B Horizon {#sec-diag-britb}

A brittle-B horizon is a B or BC horizon that has **ALL** of the following:

1.  Brittle failure (the horizon may be gravelly, but the fine-earth fraction must be sufficiently coherent to allow brittle failure); **AND**
2.  It is apedal-massive. Extremely coarse or gross prisms may be present, if the interior of the prisms is apedal-massive; **AND**
3.  Few or less fine roots occur throughout the horizon.

The brittle-B horizon differs from the [fragipan](#sec-diag-fpan) by having either lower soil strength or lower penetration resistance. Extremely coarse prism faces, if present, are not defined by low chroma colours as they are in a fragipan. The horizon commonly has some roots throughout, in contrast to the fragipan in which roots are confined to cracks. A brittle-B horizon is usually given the horizon notation suffix (x).

## Calcareous Horizon {#sec-diag-calch}

A calcareous horizon is an horizon in which calcium carbonate occurs in the fine-earth fraction. The concentration may be as low as 1%, but its presence can be detected by effervescent reaction with 10% HCl on samples from a freshly exposed profile. The calcium carbonate may be inherited from a calcareous parent material, or it may be formed in the soil and occur as coatings on stones, filamentous deposits in pores, or as nodules.

The calcareous horizon includes horizons with the more stringent requirements of the calcic horizon of Soil Taxonomy [@soilsurveystaff1999]. The less stringent limits of the calcareous horizon are needed to distinguish the calcium-carbonate-accumulating horizons in the Semiarid Soils which have developed by weathering of calcium ions from silicate minerals in non-calcareous parent materials. Calcium carbonate dusts, which have contributed to calcium carbonate enriched horizons in other countries, have not been identified in New Zealand.

## Crumb Structure {#sec-diag-crumb}

Crumb structure is defined as an earthy apedal material [@milne1991] with very friable failure class. Soil with crumb structure in situ has the gross appearance of massive soil material, but when disaggregated or disturbed, microfragments are produced with a superficial resemblance to breadcrumbs. When examined with a ×10 power hand-lens, these prove to be loosely packed aggregates of spheroidal micropeds (\<0.5 mm) with packing voids forming a prominent part of the aggregate.

## Cutanic Horizon {#sec-diag-cuth}

A cutanic horizon is a B or BC horizon containing translocated material forming dark-coloured coatings on ped faces, in pores or on coarse fragments. The coatings fail to meet the requirements for coatings of an [argillic horizon](#sec-diag-argh) or a Bh horizon.

It meets **BOTH** of the following:

1.  The coatings do not have a waxy lustre when dry or are not sufficiently thick to envelop fine sand grains. Silt coatings are excluded. (Silt coatings have similar colour to the matrix, or have higher value and/or lower chroma than the matrix. On drying they may be thick enough to meet [argillic horizon](#sec-diag-argh) requirements, and have flow-like surfaces, but they have a matt rather than a waxy lustre.) **AND**
2.  The coatings have moist colour values of 4 or less, or value 5 and chroma 3 or less. Many soils have horizons with coatings on peds or in pores which are very thin, do not have a waxy lustre when dry and have lower colour value than the matrix. It is difficult in the field to decide whether these coatings are inorganic or organic, and whether they are derived by illuviation from overlying horizons, by movement within horizons or by in-situ weathering. The [cutanic horizon](#sec-diag-cuth) is designed to recognise such horizons. A cutanic horizon is usually given the horizon notation suffix (h).

## Cutanoxidic Horizon {#sec-diag-cutoxh}

A cutanoxidic horizon [@wilson1987] is a strongly weathered, clayey, low-activity-clay horizon. The dominant clays are kaolin group minerals, and clay coatings occur on less than 10% of ped faces. Exchangeable aluminium, as a percentage of CEC, is usually greater than 25% (and is frequently more than 50% in some part of the horizon). It meets the requirements of a [cutanic horizon](#sec-diag-cuth), and has **ALL** the following characteristics:

1.  It is a B horizon that is clayey and has fine polyhedral peds; **AND**
2.  It has a failure class of friable only at water contents close to field capacity. Small changes in water content from field capacity result in large changes in soil strength and failure. Semi-deformable failure occurs at water contents wetter than field capacity. Very firm or stronger ped strength with brittle failure occurs at soil water matric potentials drier than about 30 kPa. Soil materials lack the characteristic friable failure over a wide range in moisture contents that is exhibited by [oxidic horizons](#sec-diag-oxh); **AND**
3.  Soil materials are sticky and very plastic, in comparison to oxidic materials which are only slightly sticky in relation to their clay content; **AND**
4.  Peds are larger than 2 mm and have smooth faces with silt-sized aggregates of iron oxide crystallites which give the ped faces a dusty appearance when dry. The latter property in particular distinguishes this horizon from horizons developed in well-drained Brown Soils.

The significance of this horizon lies in the combination of low ECEC, clay illuviation, acidity and "non-oxidic" physical properties.

## Densipan {#sec-diag-dens}

A densipan is a non-cemented E horizon of very high soil strength and bulk density. It meets **BOTH** of the following:

1.  Either
    (a) Unconfined strength, as measured by a resistance-to-crushing test, is hard or very hard at soil water states from near wet to dry, *or*
    (b) Soil penetration resistance measured by a 6.5 mm flat-tipped penetrometer exceeds 4000 kPa at soil water states from near wet to dry; **AND**
2.  Moist and dry samples slake in water.

Densipans occur in soils with siliceous parent materials. The strength is due to a close-fit arrangement of sand and silt-sized quartz particles. It differs from a [duripan](#sec-diag-dpan) by lack of cementation.

## Distinct Topsoil {#sec-diag-dts}

A distinct topsoil [modified from @avery1980] is normally designated an A horizon and has **BOTH** of the following:

1.  Moist colour value and/or chroma is less than that of the horizon below; **AND**
2.  Thickness is 5 cm or more (including any F, H or O topsoil layer). The distinct topsoil is used to distinguish minimal soil development in the distinction between [Recent Soils](Key-to-Orders.qmd#sec-R) and [Raw Soils](Key-to-Orders.qmd#sec-W).

## Duripan {#sec-diag-dpan}

A duripan is a subsurface horizon that is cemented by silica or other opaque or uncoloured material. It has **ALL** of the following requirements, but does **not** meet the requirements of the [calcareous horizon](#sec-diag-calch):

1.  Dry fragments do not slake in water even during prolonged wetting; **AND**
2.  It does not react visibly with 10% HCl; **AND**
3.  The average lateral distance between any fractures is 100 mm or more.

The duripan is recognised in [Pallic Soils](Key-to-Orders.qmd#sec-P) where the cementing materials are apparently related to the presence of siliceous tephra in the parent material or high exchangeable sodium in the soil.

## Fluvial Features {#sec-diag-fluv}

The intention of the term 'fluvial features' is to recognise soils with parent materials that result from transportation, sorting and deposition by water. Fluvial features are used to differentiate [Recent Soils](Key-to-Orders.qmd#sec-R), [Raw Soils](Key-to-Orders.qmd#sec-W) and parent material classes of soil series that occur on landforms formed through fluvial processes.

Relevant landforms include floodplains, estuarine surfaces, lacustrine surfaces, aggregating fan surfaces, levees, backplains, bars, channels, deltas, floodplain benches, outwash plains and swamps (all defined by @milne1991).

Confirming soil characteristics include:

1.  A buried A horizon or some other field indication of an irregular change in carbon with depth (such as sedimentary plant-leaf material).
2.  Sedimentary bedding in C horizons, indicating deposition in water (such as scoured surfaces, cross-stratification, sedimentary laminations, current ripples or foreset beds).
3.  An unripened horizon with fluid, or very fluid, fluidity class in some layer with an upper surface within 120 cm of the soil surface.
4.  In [tephric soil materials](#sec-diag-teph)
    a.  disturbance or overthickening of the regional sequence of tephras;
    b.  rounded or subrounded gravel;
    c.  presence of non-volcanic rock fragments.

The emphasis here on genetic landform criteria is not consistent with the principle that soils should be classified on the basis of similarity of measurable soil properties rather than presumed genesis (see Introduction). Measurable soil properties that will group together the required soils have not been recognised. The confirmatory soil properties, however, will aid class assignment decisions in many cases.

## Fragipan {#sec-diag-fpan}

A fragipan is an apparently non-cemented horizon that has high bulk density (usually 1.5 Mg/m^3^) with high strength when dry. It has **ALL** of the following:

1.  An air-dried clod must slake when fully immersed in water; **AND**
2.  Brittle failure when moist (the horizon may be gravelly but the fine-earth material must be Sufficiently coherent to allow brittle failure); **AND**
3.  It has at least slightly firm moist soil strength; **AND**
4.  ***Either***
    (a) Extremely coarse or gross prismatic peds: the prisms have apedal-massive interiors, or break to secondary peds with horizontal dimensions of 100 mm or more, and the prism faces are defined by colours of chroma 3 or less, or
    (b) The horizon is apedal-massive throughout, or has extremely coarse or gross prismatic peds, and the moist soil strength is very firm; **AND**
5.  If roots are present, they are confined predominantly to planar voids between prisms or to worm galleries; **AND**
6.  Moist penetration resistance measured by a 6.5 mm flat-tipped penetrometer is 3100 kPa or more; **AND**
7.  It does not occur within an E horizon.

## Gley Profile Form {#sec-diag-gleypf}

A gley profile form is defined by the presence of a [reductimorphic horizon](#sec-diag-redmh) with an upper boundary within either 15 cm of the base of the A horizon (excluding an AB or A/B horizon) or 30 cm of the mineral soil surface.

Soils with a gley profile form have usually been recognised as poorly or very poorly drained soils in the soil drainage classification of @taylor1962.

## Humus-Pan {#sec-diag-hpan}

A humus-pan is a B horizon that is 10 mm or more thick and is normally given the horizon designation Bhm.

It has **ALL** of the following requirements:

1.  It is apedal (massive); **AND**
2.  It has either firm or stronger moist soil strength, brittle failure when moist, or moist penetration resistance of 3100 kPa or more; **AND**
3.  It has dominant moist colour value in the matrix of 3 or less, or moist colour value of 4 if the chroma is 2; **AND**
4.  It contains more than 1.0% organic carbon.

## Ironstone-Pan {#sec-diag-ipan}

An ironstone-pan is an indurated horizon dominantly composed of iron oxides with or without manganese oxides. It has **ALL** of the following characteristics:

1.  The upper boundary is distinct, abrupt or sharp; **AND**
2.  It is weakly or strongly indurated; **AND**
3.  Fresh fracture surfaces are black and have a metallic lustre; **AND**
4.  It forms a continuous horizon, or it is fractured into blocks of 100 mm (in horizontal dimension) or more; **AND**
5.  It is 10 mm or more thick.

Ironstone-pans commonly occur at a textural discontinuity, in the fluctuating zone of a water-table. It is likely that the iron has been precipitated from iron-rich groundwater moving laterally. In Taranaki [@childs1990], the pans are porous and some appear to have formed as iron-oxide rhizomorphs, which have been progressively infilled and welded together by further precipitations of iron oxide. In these pans the mineralogy is dominated by varying proportions of goethite and ferrihydrite.

Ironstone-pans are not usually associated with eluvial horizons and do not occur in [Podzols](Key-to-Orders.qmd#sec-Z). They differ from [ortstein-pans](#sec-diag-opan) and [placic horizons](#sec-diag-plac) which are often associated with eluvial horizons in [Podzols](Key-to-Orders.qmd#sec-Z) or [Brown Soils](Key-to-Orders.qmd#sec-B), and which have high organic carbon levels. Ironstone pans meet the requirements of the material below a petroferric contact of Soil Taxonomy [@soilsurveystaff1999].

The pans are a barrier to plant roots. Heavy machinery is required to break them up for the installation of drains. The permeability of the pans is likely to be slow.

## Lithic Contact {#sec-diag-lithc}

A lithic contact occurs at the contact of soil with underlying rock. The rock is hard or very hard and is impracticable to dig with a spade.

*In situ* rocks in New Zealand are commonly jointed at intervals of less than 100 mm, and consequently the lithic contact definition of Soil Taxonomy [@soilsurveystaff1999] often fails to apply [@laffan1979]. The lithic contact is defined here as follows:

At a lithic contact, rock fragments accommodate one another with non-random orientation with respect to any geological structure that may be present, and cracks or joints are mostly less than 5 mm wide.

The lithic contact may be subdivided into coherent-lithic or shattered-lithic materials.

*Coherent-lithic* materials are equivalent to materials beneath the lithic contact of Soil Taxonomy. Cracks or joints occur at horizontal intervals of more than 100 mm. These materials occur on the ignimbrites, Otago schists, basalt flows of the Dunedin complex, and Fiordland granites. They often cause drainage impedance.

*Shattered-lithic* materials are similar except that joints or cracks may occur at intervals of less than 100 mm. Shattered-lithic materials differ from fragmental or skeletal materials in which there is no continuity of geological structure between adjacent rock fragments, and rock fragment faces do not accommodate one another. Shattered-lithic materials are more permeable than coherent-lithic materials, and offer a significant rooting volume.

## Mottled Profile Form {#sec-diag-mottpf}

A mottled profile form is defined by **EITHER**

1.  A [redox-mottled horizon](#sec-diag-roxh) with an upper boundary within 15 cm of the base of the A horizon, or within 30 cm of the mineral soil surface; **OR**
2.  A [reductimorphic horizon](#sec-diag-redmh) with an upper boundary between 30 and 60 cm of the mineral soil surface.

Soils with a mottled profile form have usually been recognised as imperfectly drained soils in the soil drainage classification of @taylor1962. Redox mottles are formed as a result of the reduction and solubilisation of iron and/or manganese, their translocation and concentration, and their oxidation and precipitation in the form of oxides. Mottles that have originated in some other way (e.g. rock colour patterns or skeletans) are excluded.

## Nodular Horizon {#sec-diag-nodh}

A nodular horizon has more than 15% (by volume) nodules, as segregations of iron or aluminium oxyhydroxides, with some kaolinite, in a layer more than 10 cm thick.

The nodules are common features in [Oxidic Soils](Key-to-Orders.qmd#sec-X) and some [Granular Soils](Key-to-Orders.qmd#sec-N) [@wilson1987]. The frequency distribution of nodules is clustered in the \< 2% range and the \> 15% range. Few profiles are known to lie in between.

The nodular horizon limit is intended to exclude thin layers of rewashed nodules on colluvial footslopes, and also infrequent localised concentrations in soils with characteristically few nodules.

## Organic Soil Material {#sec-diag-org}

Organic soil material is soil material dominated by organic matter, excluding fresh litter (L horizons) and living plant material. Organic soil material usually has at least 18% organic carbon (approximately 30% organic matter) but it is defined here using morphology and simple analyses for easier recognition. (For most New Zealand soils, organic carbon may be estimated by total carbon.)

Organic soil material has **EITHER**

1.  ***All*** of the following:
    (a) Colour value moist of 3 or less (after exposure to air) and colour value dry of 4 or less, **AND**
    (b) Deformable failure, **AND**
    (c) Weight loss of 65% or more by oven-drying a field-saturated sample; **OR**
2.  More than 20% (by volume) unrubbed fibre content; **OR**
3.  More than 35% (by weight) loss on ignition except in materials dominated by allophanic soil material or by limestone; **OR**
4.  18% or more total carbon.

Organic soil materials that have been accumulated under wet conditions are subdivided into three classes, based on evidence of decomposition [@clayden1989]. These classes are used to distinguish soil groups of [Organic Soils](Key-to-Orders.qmd#sec-O).

*Fibric soil material* (Of horizon) consists mainly of well preserved plant remains that are readily identifiable in terms of botanical origin. The fibre content after rubbing is at least 75% by volume.

Fibres are pieces of plant tissue large enough to be retained on a 100-mesh (0.15 mm) sieve, except for wood fragments that cannot be crushed or shredded in the hand and are larger than 2 cm in the smallest dimension. Rubbed fibre is the fibre that remains after rubbing a wet sample 10 times between the thumb and forefinger, or kneading a ball in the palm 10 times using firm pressure.

*Mesic soil material* (Om horizon) consists mainly of partially decomposed plant remains (semi-fibrous peat or hemic soil material) and does not meet the requirements of either fibric soil material or humified soil material.

*Humified soil material* (Oh horizon) consists of strongly decomposed organic material (humified peat or sapric soil material) with few or no identifiable plant remains other than resistant woody fragments \> 20 mm that cannot be reduced to fibres by crushing and shredding between the fingers. The fibre content is less than 15% after rubbing.

## Ortstein-Pan {#sec-diag-opan}

An ortstein-pan is a B horizon that is normally given the horizon notation Bsm. It has **ALL** of the following requirements:

1.  Thickness of more than 10 mm; **AND**
2.  The upper boundary is sharp or abrupt; **AND**
3.  It is massive and has either firm or stronger moist soil strength, or has moist penetration resistance of 3100 kPa or more; **AND**
4.  It does not meet the requirements of a [humus-pan](#sec-diag-hpan), and does not have the metallic lustre of fresh fracture surfaces of an [ironstone-pan](#sec-diag-ipan).

## Oxidic Horizon {#sec-diag-oxh}

The oxidic horizon is a strongly weathered B horizon consisting of mixed crystalline iron and aluminium oxides and kaolin group minerals, with low activity clay properties. It has **ALL** the following requirements:

1.  Weak or very weak primary ped strength and soil strength as determined by the unconfined resistance-to-crushing test at moist to dry soil water states; **AND**
2.  Unconfined failure is friable or very friable over very moist to dry soil water states. Materials fail to predominantly 3 mm or smaller peds comprising silt- and sand-sized polyhedra and spheroids; **AND**
3.  Primary peds slake rapidly in water to stable microaggregates which show no dispersion or slight dispersion after 100 inversions using the method of @mcqueen1981; **AND**
4.  Non-reactive or very weakly reactive to the reactive-aluminium test [@fieldes1966]. Oxidic horizons are clayey, with measured clay contents commonly exceeding 60%. The measured clay percentage is usually larger than in overlying A horizons, but clay increase is not a defining criterion because of the problem of quantifying clay contents in materials that are frequently difficult to disperse.

Materials are only slightly sticky and plastic in relation to clay content. Clay coatings are either visually absent or only present at frequencies of about 1%. The oxidic horizon has low activity clay accessory properties with ECEC and CEC less than 12 and 16 cmol/100g clay respectively.

## Paralithic Contact {#sec-diag-plith}

A paralithic contact is the upper surface of rock or regolith material that has **ALL** of the following requirements:

1.  It can be cut with difficulty with a spade; **AND**
2.  Wet penetration resistance exceeds 2600 kPa; **AND**
3.  Roots if present are few and confined to cracks; **AND**
4.  If the overlying horizon is a [reductimorphic](#sec-diag-redmh) or [redox-mottled](#sec-diag-roxh) horizon, low chroma or high chroma mottles are less common below the contact.

The paralithic contact meets the definition of a paralithic contact of Soil Taxonomy [@soilsurveystaff1999], but without the restrictive requirement for spacing of cracks. The horizon beneath the contact is given the horizon designation CR. Paralithic contacts may occur either on weakly weathered or unweathered rocks which are not strongly lithified, or on saprolites which have become soft by strong weathering.

## Peaty Topsoil {#sec-diag-pts}

A peaty topsoil is 10 cm or more thick and is saturated for 30 or more consecutive days in most years (unless it is artificially drained), and has **EITHER**

1.  Peat, sandy peat or loamy peat texture, **OR**
2.  Slightly peaty texture (17--30% organic matter) if the clay content is less than 18%.

In some subgroups a peaty topsoil may be buried by a surface mantle of new material of up to 60 cm in thickness.

## Perch-Gley Features {#sec-diag-pgley}

Perch-gley features are the morphologic indicators of saturation and reducing conditions caused by a water-table perched on a [slowly permeable layer](#sec-diag-slowp) within the soil profile.

A horizon with perch-gley features **EITHER**

1.  Has [redox-segregations](#sec-diag-rsegs) that occur mainly within peds, or in the case of an apedal soil, mainly within the soil matrix. Macro-void surfaces, either partings or pores, are dominated by greyish colours (moist chroma 2 or less, or moist chroma 3 and value 6 or more). Iron and manganese precipitates occur either adjacent to the greyish void surfaces as a selvedge or as discrete mottles within the soil mass; **OR**
2.  Overlies a horizon that is less gleyed (e.g. less redox-segregations) with a matrix that is not dominated by greyish colours.

## Placic Horizon {#sec-diag-plac}

A placic horizon is a thin iron pan that is normally designated Bfm. It has **ALL** of the following:

1.  It is 10 mm or less thick; **AND**
2.  It is at least weakly indurated, and is black to reddish brown or dark red in colour. A black upper part can often be distinguished from a reddish brown lower part; **AND**
3.  The upper and lower boundaries are sharp, and may be smooth, wavy or convolute in shape.

The placic horizon usually occurs as a single pan but in places can be bifurcated. It is equivalent to the placic horizon of Soil Taxonomy [@soilsurveystaff1999] except that New Zealand iron pans are enriched in iron and organic matter without significant accumulations of manganese [@clayden1990].

## Podzolic-B Horizon {#sec-diag-podzb}

A podzolic-B horizon is a B horizon that meets one of the following:

1.  It meets the requirement of a Bh horizon (because it has colour value and chroma of 3 or less, or value 4 and chroma 2, dominant in the matrix, and contains more than 1% organic carbon). The fabric has sand- or silt-size pellet-like aggregates, coats on mineral grains, or both; **OR**
2.  
    (a) It is associated with an overlying (but not necessarily immediately overlying) E horizon (i.e. Ea horizon) in which weathered films on sand and silt particles are either absent, very thin or discontinuous, so that the colour of the horizon is mainly determined by the colours of the uncoated grains. The moist colour value is 4 or more (or a dry colour value is 5 or more). It has higher colour value or lower chroma and less well developed pedality than an underlying B horizon; **AND**
    (b) The B (or some subhorizon of the B) horizon is 5 cm or more thick and meets the requirements of a Bs (or Bs(g) or Bs(f)) horizon because it has a strong or very strong reactive-aluminium test, and at least one of the following:
        (i) Reddish hue and highest chroma at the top of the horizon; *or*
        (ii) Earthy apedal with fine spheroids, or weakly developed blocks or polyhedra; *or*
        (iii) Very weak or weak soil strength when moist or dry; *or*
        (iv) Sand- or silt-sized pellet-like aggregates; **OR**
3.  It meets the definition of a Bs horizon (part 2(b) above) and has in addition, coatings of value 4 or less ***either***
    (a) On 50% or more ped faces, *or*
    (b) As patches covering 20% or more of cut faces.

## Reactive-Aluminium Test {#sec-diag-naf}

This test indicates the presence of reactive hydroxy-aluminium groups, as occur for example in allophane and aluminium-humus complexes [@milne1991].

Using the procedure of @fieldes1966, 1 drop of saturated sodium fluoride (NaF) solution is placed on a small test sample of soil, which has been smeared on to a filter paper treated with phenolphthalein indicator. The soil sample must be field moist. For classification, the reactivity of the soil sample is placed into one of the following classes.

```{r}
table_003 <- data.frame(
  'Code' = seq(0,5,1),
  'Name' = c('Non-reactive',
              'Very Weak',
              'Weak',
              'Moderate',
              'Strong',
              'Very Strong'),
  'Definition' = c('No colour within 2 minutes',
                   'Pale red or light red (5R 6/1) just discernible within 2 minutes.',
                   'Pale red or light red (5R 6/1) within 1 minute.',
                   'Red or weak red (5R 4 or 5/-) within 1 minute.',
                   'Dusky red or dark red (5R 3/-) after 10 seconds.',
                   'Dusky red or dark red (5R 3/-) within 10 seconds')
  )

ft <- flextable(table_003)
ft <- bold(ft, part = 'head')
ft <- set_caption(ft, 'Reactive-aluminium test interpretation.')
ft <- set_table_properties(ft, layout = "autofit", width = 1)
ft
```

## Redox-Mottled Horizon {#sec-diag-roxh}

A redox-mottled horizon is a horizon affected in parts by reducing conditions as indicated by the presence of [redox-segregations](#sec-diag-rsegs). These usually indicate intermittent saturation of the soil by water.

A redox-mottled horizon has 2% or more redox segregations. If low chroma colours (moist chroma 2 or less, or moist chroma 3 with value 6 or more) occur, they must occupy less than 50% of the matrix exposed in a cut face of the horizon and are not dominant on ped faces.

The intermittent wetness may be caused by intermittent perched water, or by the fluctuating upper limits of deeper more prolonged groundwater. A redox-mottled horizon may represent more prolonged saturation and reduction in parent materials that are predominantly andesitic or basaltic compared with other parent materials.

## Redox Segregations {#sec-diag-rsegs}

Redox segregations are mottles or concretions formed as a result of the reduction and solubilisation of iron and/or manganese, their translocation, concentration, and their oxidation and precipitation in the form of oxides @clayden1989. They may occur as low or high chroma colours, or both.

The nature of the water table is indicated by the association of low and high chroma colours. If subject to reduction by perched water, the low chroma colours are likely to be at ped or pore surfaces and the high chroma colours are likely to be within the soil matrix. If the soil is subject to reduction by groundwater, the low chroma colours are likely to be within the soil matrix and the high chroma colours are likely to be at ped or pore surfaces. Reducing conditions may also be indicated by the presence of sufficient ferrous iron to give a positive reaction to α,α'-dipyridyl [@childs1981].

## Reductimorphic Horizon {#sec-diag-redmh}

A reductimorphic horizon has a slightly peaty texture class, or low chroma colours (moist chroma 2 or less, or moist chroma 3 with value 6 or more) that occupy 50% or more of the matrix exposed in a cut face of the horizon or are dominant on ped faces. A reductimorphic horizon includes any subjacent layers or interlayers of peaty soil material.

A reductimorphic horizon is a horizon strongly affected by reducing conditions as indicated by greyish colours consistent with long saturation by water. The prolonged wetness may be caused by a water-table perched on a slowly permeable layer within the soil profile or by a groundwater-table.

## Slowly Permeable Layer {#sec-diag-slowp}

A slowly permeable layer is one in which the vertical saturated hydraulic conductivity is less than 4 mm/h (1.0 × 10^--6^ m/s) as measured by a standard method. If no measurement is available then the horizon can be identified by the following morphological characteristics [@griffiths1991]:

A slowly permeable layer meets **EITHER**

1.  

    (a) The soil material is pedal; **AND**
    (b) More than half of the peds are coarser than 10 mm (mean of the x and y axes in a horizontal plane) and meet one of the following:
        (i) Peds 20 to 50 mm, with degree of packing at least extremely high; *or*
        (ii) Peds 50 to 100 mm with degree of packing at least very high; *or*
        (iii) Peds 100+ mm with degree of packing at least high; **OR**

2.  ***Either***

    (a) The soil material is sand or loamy sand and has an extremely high degree of packing; *or*
    (b) The soil material has a particle-size group other than sandy and has at least a high degree of packing.

A slowly permeable layer is significant for the land use, genetic and hydrological understanding of soils. It may overlap a number of other diagnostic horizons --- for example, the [fragipan](#sec-diag-fpan), [argillic horizon](#sec-diag-argh), [densipan](#sec-diag-dens), [humus-pan](#sec-diag-hpan) or [ortstein](#sec-diag-opan) horizons and [lithic](#sec-diag-lithc) or [paralithic contacts](#sec-diag-plith).

## Sodic Features {#sec-diag-sod}

A horizon with sodic features has significant exchangeable sodium (not necessarily related to a high soluble salt content). It has **BOTH**

1.  **Either**
    (a) Exchangeable sodium percentage of 6% or more (or exchangeable sodium is 0.7 cmol/kg or more); *or*
    (b) When a 10 mm diameter sample which has been air-dried is placed in distilled water or salt free water a cloud of dispersed clay will form within 10 minutes around the sample. This test will not apply if the soil is in any degree cemented; **AND**
2.  **Either**
    (a) Clay or clay/organic coatings have colour value 4 or less; *or*
    (b) Prismatic or blocky peds; *or*
    (c) It may be overlain by an Ew, Ew(g) or Ew(f) horizon, or have skeletans (visible on dry ped faces) near the top of the horizon.

## Tephric Soil Material {#sec-diag-teph}

Tephric soil material occurs in or below the soil solum. It includes:

1.  Tephra --- unconsolidated, primary pyroclastic products of volcanic eruptions [@froggatt1990] (including ash, cinders, lapilli, pumice, pumice-like vesicular pyroclastics, blocks, or volcanic bombs), and
2.  Tephra deposits --- material derived at least partly from tephra that has been reworked and mixed with material from other sources. They include tephric loess, tephric blown sand and volcanogenic alluvium. As a general guide, tephric deposits from andesitic sources have more than 10% volcanic glass in the sand fraction and those from rhyolitic sources have more than 40% volcanic glass in the sand fraction.

Tephric soil material may include soil materials that meet the requirements of [allophanic soil material](#sec-diag-alloph) or [vitric soil material](#sec-diag-vitr). It is used to distinguish soil groups of the [Raw Soils](Key-to-Orders.qmd#sec-W) and [Recent Soils](Key-to-Orders.qmd#sec-R), and parent material classes at soil series level.

## Vitric Soil Material {#sec-diag-vitr}

Vitric soil material [@parfitt1984] has more than 35% coarse-fraction (2 mm or greater, by volume) of which 60% is pumice or cinders, or there is more than 40% sand of which more than 30% is volcanic glass (or crystals coated with glass) [@eden1992].

## Weathered-B Horizon {#sec-diag-bw}

A weathered-B horizon shows evidence of alteration and is normally designated Bw, Bw(g), Bw(f), etc. It has at least **ONE** of the following:

1.  Redder hue or higher chroma than an underlying horizon in similar materials; **OR**
2.  Have spheroidal, blocky, polyhedral, tabular, prismatic, columnar or platy pedality which distinguish the horizon from a BC or C horizon below; **OR**
3.  Evidence of either partial or complete decalcification, i.e. less CaCO~3~ than the underlying horizon which may contain redeposited carbonates.

A weathered-B horizon may also meet the requirements of a [redox-mottled horizon](#sec-diag-roxh), [argillic horizon](#sec-diag-argh), [cutanic horizon](#sec-diag-cuth), or [brittle-B horizon](#sec-diag-britb).
